---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# igr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/igr)](https://CRAN.R-project.org/package=igr)
<!-- badges: end -->

Convert Irish Grid References to Irish Grid coordinates or a simple feature object in any coordinate reference system, and vice versa.

## Installation

You can install the development version of igr like so:

``` r
# FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE?
```

## Example

### Converting from Irish Grid References

Convert Irish Grid References to Irish Grid coordinates:

```{r example-igr}
library(igr)

igrs <- c("A09", "B", "D12", "N8090", "S1234588800", "Z90")

igr_to_ig(igrs)

```

Convert Irish Grid References to an sf object with Irish Grid coordinate reference system:

```{r example-igr-sf}
p_sf <- st_igr_as_sf(data.frame(igr = igrs), "igr")

p_sf

```

Plot the results:

```{r example-igr-sf-plot, message=FALSE}
library(maps)
library(tmap)

ie_uk <- maps::map("world", regions = c("Ireland", "UK"), plot = FALSE, fill = TRUE)
ie_uk_sf <- sf::st_as_sf(ie_uk)

tm_shape(p_sf, ext = 1.2) +
  tm_dots(size = 1, col = "cornflowerblue") +
  tm_text("igr", ymod = 1) +
  tm_shape(ie_uk_sf) +
  tm_borders()
```

### Converting to Irish Grid References

Starting with a list of Irish Grid coordinates:

```{r example-igc-igr}

p <- list(x = c(0, 490000), y = c(400000, 0))

ig_to_igr(p)
ig_to_igr(p, sep = " ")
ig_to_igr(p, digits = 1)

```


Starting with an sf object:

```{r example-sf-igr}

p_sf <- sf::st_as_sf(data.frame(p),
 crs = 29903,
 coords = c("x", "y")
)

st_irishgridrefs(p_sf)
st_irishgridrefs(p_sf, sep = " ")
st_irishgridrefs(p_sf, digits = 1)

```


