---
title: "Irish Grid References in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Irish Grid References in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

plot_maps <- requireNamespace("maps", quietly = TRUE) & requireNamespace("tmap", quietly = TRUE)
```

## Introduction

The Irish Grid ([EPSG:29903](https://epsg.io/29903)) spans the island of Ireland and was defined by the Ordnance Survey Ireland (now [Tailte Éireann](https://www.tailte.ie)) and [Ordnance Survey of Northern Ireland](https://www.nidirect.gov.uk/campaigns/ordnance-survey-of-northern-ireland). Locations within the Irish Grid are defined by an X and Y coordinate.

Irish grid references are a way of recording locations on the Irish Grid, and were particularly relevant when dealing with paper-based maps. An Irish grid reference consists of a letter, optionally followed by an easting and northing. A letter alone (e.g. "N") refers to a particular 100 km square on the Irish Grid. An Irish grid reference with a 1-digit easting and northing (e.g. "N85") refers to a 10 km square. A 5-digit easting and northing (e.g. "N 12345 67890") refers to a 1 m square.

There are many R packages facilitating [spatial data analysis and visualisation](https://cran.r-project.org/web/views/Spatial.html). The igr package is designed to simplify using datasets containing Irish grid references in R. It translates Irish grid references into Irish Grid coordinates or sf objects that other R packages can work with. The igr package supports precisions of Irish grid reference in multiples of ten from 1 m to 100 km, mixes of precision, and Irish grid references with or without whitespace between letter, easting and northing.

When converting Irish grid references to point locations (using `igr_to_ig()`, or `st_igr_as_sf()` with the default `polygons=FALSE`), the south west corners of the relevant Irish Grid squares are returned. When converting to polygons (using `st_igr_as_sf()` with `polygons=TRUE`), the polygons returned span the entire square of each grid reference.

The functions `st_igr_as_sf()` and `st_irishgridrefs()` convert to and from [sf](https://r-spatial.github.io/sf/) (simple feature) objects in any coordinate reference system.

## Installation

The development version of igr can be installed from github as follows:

```{r install, eval=FALSE}
# Install devtools package if needed
install.packages("devtools")

# Load devtools package if needed
library(devtools)

# Install development version of igr package from github
install_github("digitalnature-ie/igr")
```

## Loading the package

The igr package is loaded in the normal way:

```{r load}
library(igr)
```

## Converting from Irish grid references

To convert a vector of Irish grid references to Irish Grid coordinates use `igr_to_ig()`.

```{r example-igr-1}
igrs <- c("A", "D12", "J53", "M5090", "N876274", "S1234550000", "W")

igr_to_ig(igrs)
```

The Irish grid references may have spaces between the grid letter, easting and northing:

```{r example-igr-2}
igrs <- c("A", "D 12", "J 5 3", "M 50 90", "N 876 274", "S 12345 50000", "W")

igr_to_ig(igrs)
```

These sample Irish grid references have a mix of precisions. The precisions in metres can be returned by supplying a column name for the precision data:

```{r example-igr-res}
igr_to_ig(igrs, precision = "prec")
```

A list or data.frame of Irish grid references can be converted to an sf object using `st_igr_as_sf()`. By default the geometries returned are points located at the south west corner of each grid reference, and the coordinate reference system is Irish Grid ([EPSG:29903](https://epsg.io/29903)):

```{r example-igr-poi}
igr_df <- data.frame(igr = igrs)

points_sf <- st_igr_as_sf(igr_df, "igr")

points_sf
```

These points can be plotted by the [tmap](https://r-tmap.github.io/tmap/) package. First a basic outline of Ireland and the United Kingdom can be retrieved from the [maps](https://cran.r-project.org/package=maps) package.

```{r example-igr-poi-plot, fig.height=4, message=FALSE, eval=plot_maps}
library(maps)
library(tmap)

ie_uk_sf <- maps::map("world",
  regions = c("Ireland", "UK"),
  plot = FALSE,
  fill = TRUE
) |>
  sf::st_as_sf(ie_uk) |>
  sf::st_transform(29903)

tm_shape(points_sf, ext = 1.4) +
  tm_dots(size = 1, col = "cornflowerblue") +
  tm_text("igr", ymod = 1) +
  tm_shape(ie_uk_sf) +
  tm_borders()
```

The Irish grid references can also be converted to polygon geometries. Each polygon spans the full extent of a grid reference, so if grid references with a mix of precisions are supplied, the polygons will have a mix of sizes.

```{r example-igr-pol}
polygons_sf <- st_igr_as_sf(igr_df, "igr", polygons = TRUE)

polygons_sf
```

Plotting these polygons:

```{r  example-igr-pol-plot, fig.height=4, eval=plot_maps}
tm_shape(ie_uk_sf, bbox = points_sf, ext = 1.4) +
  tm_borders() +
  tm_shape(polygons_sf, ext = 1.2) +
  tm_polygons(size = 1, col = "cornflowerblue", alpha = 0.5) +
  tm_text("igr", ymod = -1)
```

## Converting to Irish grid references

Irish Grid coordinates can be converted to Irish grid references using `ig_to_igr()`. The first two columns in the matrix supplied must be the Irish Grid X coordinate and Y coordinate respectively. Any other columns are ignored.

```{r example-ig-1}
p <- matrix(c(0, 490000, 400000, 0, 453000, 4000), ncol = 2, byrow = TRUE)
colnames(p) <- c("x", "y")

p
ig_to_igr(p)
```

Spaces or other strings can be inserted between the letter, easting and northing:

```{r example-ig-2}
ig_to_igr(p, sep = " ")
```

The Irish grid references can be limited to a particular precision by specifying the required precision in metres. Multiples of 10 from 1 m to 100,000 m (100 km) are supported:

```{r example-ig-3}
ig_to_igr(p, precision = 10000) # 10 km precision

ig_to_igr(p, precision = 1) # 1 m precision
```

A data.frame of Irish Grid coordinates can be converted:

```{r example-ig-4}
p_df <- data.frame(p)

p_df
ig_to_igr(p_df)
```

An sf object containing geometries of type POINT can be converted to Irish grid references using `st_irishgridrefs()`:

```{r example-ig-5}
p_sf <- sf::st_as_sf(p_df,
  crs = 29903,
  coords = c("x", "y")
)

p_sf
st_irishgridrefs(p_sf)
```

To append Irish grid references (with spaces) to the original sf object using base r:

```{r example-ig-6}
p_sf$igr <- st_irishgridrefs(p_sf, sep = " ")

p_sf
```

In tidy r this can be written as:

```{r example-ig-7, eval = FALSE}
p_sf <- p_sf |>
  dplyr::mutate(igr = st_irishgridrefs(sep = " "))
```
